Is Information Ontological or Physical, or Is It Perhaps Something Else? Some Remarks on Krzanowski's Approach to the Concept of Information

### 2023-02-23 ART_Martin/Martin.tex
Editorial corrections added

\myquote{
 Present day computers are very different from the early stand-alone calculators designed for mechanizing mathematical operations. They are largely \emph{used for communication in world-wide networks} and variety of information processing and knowledge management. Moreover, they play \emph{an important role in the control of physical processes and thus connect to the physical world}, especially in automation and robotics. [\ldots] Computational processes are nowadays \emph{distributed, reactive, agent-based, and concurrent}. The main criterion of success of the computation is not its termination, but its \emph{response to the outside world}, its speed, generality and flexibility; adaptability, and tolerance to noise, error, faults, and damage. \parencite{Dodig-Crnkovic2011}
 }

@@

\myquote{
A~fairly obvious, trivial, and almost-circular definition of ‘computer’ says that a~computer is a~machine that computes. The natural next question is: What does it mean to compute? But this shifts the burden of answering our question away from what computers are to the topic of what computation is. Many of the objections to various theories about computers are really objections to what counts as a~computation. \parencite{Rapaport2018}. 
}

@@

To tackle the issue of interactive computing devices, we propose here an approach that, to the best of our knowledge, has not been proposed so far: we want to examine the models of computation proposed in theoretical computer science to think about interactive computing systems. We offer a~literature survey where one can find \textit{explicit theories of interaction} \footnote{We insist on our two criteria: \textit{explicit theories of interaction} in \textit{theoretical computer science}. We have in mind the fact that other communities e.g., the engineering community on reactive systems, are related to our topic but they have not conceptualized \textit{interaction} as such.}. We show that the formal modelling of interactive computing systems has been brought down to whether the new interaction models are reducible to Turing’s \emph{a-machines} \parencite{Turing1937}---we will refer to them as Turing Machines (TMs). 
Questioning the theoretical bounds of the Turing Machine in computer science when faced with the existence of interactive computing devices has been explored at least since Milner's work on communicating and mobile systems \parencite{Milner1993, Milner1999}. To the best of our knowledge, there are three areas where interaction models are framed as such. These areas are some works (i) on concurrency by Milner and his followers \parencite{Milner1999, Milner2006}, (ii) on Reactive Turing Machines \parencite{Andersen1997, Baeten2013, VanLeeuwen2001, VanLeeuwen2006}, and (iii) on interaction as a~new non-algorithmic computing paradigm \parencite{Goldin2006, Wegner1997, Wegner2003}. For each of the three identified models, we:

@@

Milner introduced the opposition between \textit{interactional} and \textit{computational} behaviour. Introducing the concept of interaction, Milner \parencite*{Milner1975, Milner1982, Milner1983} referred to concurrent message passing between agents. Milner's work coincided with Petri’s \parencite*{Petri1980} new model of concurrent processes, which generally intended to describe concurrency in information systems.\footnote{Concurrency theory emerged from Dick Karp's early work in the 1960s, grew with \parencite{Petri1980} and later work on transition systems \parencite{Nielsen1981, Glabbeek2004},
and has now developed into a~mature theory of reactive systems \parencite{Harel1985} with diverse network models \parencites[for an overview, see][]{Lee1998}{Lee2006_modelsofcomputation}}
 To Milner, interaction is more \textit{expressive} than a~TM, but it still describes an \textit{effective} procedure. Milner did not assert \textit{equivalence} between an interactive model and a~TM, but he introduced the topic \parencite{Milner1999} and left it unanswered. 
Four main differences between old (computational) and new (interactional) computing are made striking by Milner. First, in Milner’s words, a~Turing Machine prescribes a~behaviour to be executed. In contrast, new computing requires the description of an information flow between several system components. Second, old computing is characterized by a~\textit{hierarchical design} when current practice involves \textit{heterarchical} phenomena in the computing system. Third, in new computing, the designer cannot predict when agents will be triggered or the overall behaviour of the computing system. Fourth, the user is not merely looking for an end result in new computing practice. There is more than a~mathematical function to evaluate, as it used to be in old computing. The user instead interacts with the system, and the look for an end result is replaced by continuing interaction. Having taken stock of the evolution of computing practice on the engineering side, Milner examines its consequences on the logic foundations of computing. The pi-calculus and his work on the equivalence with automata, known as bisimulation, achieved this reflection on interactive processes \parencite{Milner1993, Milner1999} with a~formalism.

@@

If we look for a~mechanistic explanation of computing, we need some elements to be unpacked beyond a~formal account to make sense of the quoted scenario above. For example, we need to account for how reading and writing on the ports are possible. It presupposes that the interactive computing system can wait, pause, and react depending on the arrival or absence of new data. What allows such behavior? It presupposes some mechanisms allowing the system either to be interrupted by environmental processes or to check the new incoming values steadily.\footnote{More on these mechanisms and on the limitations of oracles can be found in \parencite{Martin2023}} 
In other words, given the initial question (``what is an interactive computer?''), some phenomena cannot be accounted for within the frame of an extended Turing machine. The way oracles work remains at a~level of abstraction too remote from the minimal causal blueprint we need for our purpose.

@@

\begin{table}[]
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}llll@{}}
\rowcolor[HTML]{ECF4FF} 
%PP \toprule
\multicolumn{1}{c}{} &
  \multicolumn{1}{c}{\textbf{\begin{tabular}[c]{@{}c@{}}\large Interaction as concurrent \\ \large communicating systems\end{tabular}}} &
  \multicolumn{1}{c}{\textbf{\begin{tabular}[c]{@{}c@{}}\large Interaction as extended\\ \large Turing Machines\end{tabular}}} &
  \multicolumn{1}{c}{\textbf{\begin{tabular}[c]{@{}c@{}}\large Interaction as a~new\\ \large paradigm\end{tabular}}} \\ 
%PP  \midrule
\rowcolor[HTML]{FFFFFF} 
\textbf{\large Motivation} &
  \begin{tabular}[c]{@{}l@{}}Provides new logical\\ foundations to fit new\\ engineering challenges,\\ especially concurrency\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Extends the TM\\ model to account for\\ interactive devices\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Debunks the strong Church-\\ Turing thesis\\ \\ Discusses the scope of \\ algorithmic solving\\ \\ Prones the need for a~new\\ computing paradigm\end{tabular} \\
\rowcolor[HTML]{CBCEFB} 
\textbf{\begin{tabular}[c]{@{}l@{}}\large Account for\\ \large interaction\end{tabular}} &
  \begin{tabular}[c]{@{}l@{}}Information flow\\ \\ Heterarchical design\\ \\ No complete prediction about\\ overall behavior\\ \\ No end-result\\ \\ Process calculi\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}External data needed\\ during computation\\ \\ Non-uniformity of programs\\ \\ Interaction with agents\\ \\ Infinity of operations\\ \\ Interactive machines are TMs\\ with advice\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Computers have rich interaction\\ with the environment during\\ computing execution, \\ but this processing\\ is not merely algorithmic\end{tabular} \\
\rowcolor[HTML]{FFFFFF}  
\textbf{\begin{tabular}[c]{@{}l@{}}\large Uses and\\ \large criticisms\end{tabular}} &
  \begin{tabular}[c]{@{}l@{}}First conceptualization of \\ interaction\\ \\ Legacy for automata\\ theory\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Inspires the need for a~new\\ paradigm\\ \\ Puts at the forefront the\\ Church-Turing thesis\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Controversy about \\ the powerfulness of\\ the TM\end{tabular} \\
\rowcolor[HTML]{CBCEFB} 
\textbf{\begin{tabular}[c]{@{}l@{}} \large Issues for an account \\ \large of interactive computing\end{tabular}} &
  \begin{tabular}[c]{@{}l@{}}Definition of interaction\\ restricted to specific\\ properties: concurrency and\\ communication\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Formal oracles cannot\\ account for the physical \\ possibility of entering new\\ data\end{tabular} &
  \begin{tabular}[c]{@{}l@{}} Issues about powerfulness \\ and expressiveness constrict the\\ debate in the realm of\\ computability theory\end{tabular} \\ 
%PP \bottomrule
\end{tabular}%
}
\caption{Sum-up: an overview of explicit theories of interactive computing systems in theoretical computer science}
\label{tab:my-table}
\end{table}

@@

But in a~narrow sense of expressiveness, one can interpret (wrongly, we think) the possibility of a~new paradigm as follows. Wegner and the tenants of Reactive Turing Machines could think of their interactive model as more expressive than a~TM, allowing their model to execute \textit{more functions}, even some of them being uncomputable functions in the sense of the Church-Turing thesis, solving the halting problem. In that case, the claim would indeed be controversial. The bold claim would be the following: a~TM is not only providing an account for algorithmic problem-solving through effective procedures but it could also be extended to account for other non-algorithmic processes, solving the uncomputable. Interaction would be some super-calculus, extending the calculative power of the original TM to account for interaction. It would be satisfactorily modeled with a~TM, only given more calculation power. It would go down the track of Accelerating Machines or Super-Turing Machines, able to calculate more than Turing's computable functions \cite{Copeland2002, Copeland2011, MacLennan2009}.

@@

Such lessons have just started to be drawn. They have motivated, for example, distinctions between computational models and computational explanations \parencite{Klein2020} or between models of computation and computational mechanisms \parencite{Mikowski2014}). The lesson drawn is that formal models of computing systems do not provide us with the appropriate and complete level of description to build an explanation, which is expected to identify the relevant mechanisms at stake. More precisely, an explanation for computing phenomena requires bridging a~high-level description of a~computation and its blueprint \parencite{Mikowski2011, Mikowski2016}.
The approach is based on the standard of mechanistic explanation in science, coupled with the idea that a~computational process is intrinsically mechanistic:

\myquote{
Computational explanations, according to the mechanistic account are constitutive mechanistic explanations: they explain how a~mechanism's computational capacity is generated by the orchestrated operation of its component parts. To say that a~mechanism implements a~computation is to claim that the causal organization of the mechanism is such that the input and output information streams are causally linked and that this link, along with the specific structure of information processing, is completely described. \parencite{Mikowski2014}.
}

@@

\myquote{
There are two ways in which computational models may correspond to mechanisms: first, they may be \textit{weakly equivalent} to the explanandum phenomenon, in that they only describe the input and output information, or \textit{strongly equivalent}, when they also correspond to the process that generates the output information. \parencite{Mikowski2016}
}

@@

\myquote{
Turing machines were not invented to be implemented physically at all, but some people still build them for fun. [\ldots] Imagine a~physical instantiation of a~trivial logical negation Turing machine, built of, say, steel and rubber and printing symbols on paper tape. Its alphabet of symbols consists of ``F'' and ``T.'' If the machine finds ``T'' on its tape, it rewrites it to ``F'' and halts; if it finds ``F,'' it rewrites it to ``T'' and halts. Let us suppose that the machine’s head is so old and worn out that it tears the paper tape during the readout. As a~result, no symbol will appear. [\ldots] Only when we describe the Turing machine literally, as a~causal system that has a~particular causal blueprint (engineering specifications of how it is built), can we causally predict such a~breakdown. [\ldots] Why are breakdowns and malfunctions so important? They help us discover the causal complexity of the system. [\ldots] an abstract model of computation will not predict all the possible outcomes of the breakdown, as it abstracts away from a~number of the system’s causal characteristics. So it will not tell us what is going to happen with the head; it will only say that the computation will no longer be correct. \parencite{Mikowski2011}
}

@@

We started from the need to update a~question in the philosophy of computing: \textit{what is a~computer}? Today's computers are highly interactive, so the question can be rephrased more precisely: \textit{what is an interactive computing system}? It is common to understand computers in terms of existing models of computation, hypothesizing that a~computer is primarily a~machine that carries out computation. Therefore, the working hypothesis has traditionally answered the initial question by asking what computation is. As already noted, this shift should not be taken for granted. There are, however, and the length of the paper does not allow it, historical and epistemological reasons for this shift that have been described and discussed \parencite{Daylight2014, DeMol2018a, Haigh2020}. We have chosen in this paper to ask ourselves if the shift is relevant in the case of interactive computing: do we understand what an interactive computer is by questioning the formal models proposed in theoretical computer science for interaction? Our literature review shows that there are better paths. There are two reasons for this. First, the conceptualization of interactive systems in theoretical computer science has focused on their comparison with the Turing machine (and sometimes other classical models), putting forward formal questions about powerfulness and equivalence of models that do not clarify the singularity of interactive systems from an epistemic point of view (rather than formal). There is an inherent difficulty in looking for an explanation of a~computing phenomenon in a~formal model: it needs more bricks to describe the mechanisms at stake, at a~level of abstraction operating the junction between a~high-level formalism and the blueprint. This work does not lead us to an aporia but to a~research program in the philosophy of computing: we must produce the right level of explanation for interactive computing.\footnote{More considerations on such a~research program are fleshed out in a~forthcoming paper \parencite{Martin2023}} This implies an identification of the mechanisms at play that make possible the interaction between processes within the computing machine (whether there are to be thought of as physical or computational processes, or a~mix of both\footnote{The distinction between computational and physical processes is out of the scope of this paper but more on this can be found e.g., in \parencite{Kycia_Niemczynowicz_2020}.}) and the environment. The components of such a~mechanism are to be identified and described at a~level of abstraction that allows a~satisfactory reference to the implementation.

>>>>

\myquote{
 Present day computers are very different from the early stand-alone calculators designed for mechanizing mathematical operations. They are largely \emph{used for communication in world-wide networks} and variety of information processing and knowledge management. Moreover, they play \emph{an important role in the control of physical processes and thus connect to the physical world}, especially in automation and robotics. [\ldots] Computational processes are nowadays \emph{distributed, reactive, agent-based, and concurrent}. The main criterion of success of the computation is not its termination, but its \emph{response to the outside world}, its speed, generality and flexibility; adaptability, and tolerance to noise, error, faults, and damage \parencite{Dodig-Crnkovic2011}.
 }

@@

\myquote{
A~fairly obvious, trivial, and almost-circular definition of ‘computer’ says that a~computer is a~machine that computes. The natural next question is: What does it mean to compute? But this shifts the burden of answering our question away from what computers are to the topic of what computation is. Many of the objections to various theories about computers are really objections to what counts as a~computation \parencite{Rapaport2018}. 
}

@@

To tackle the issue of interactive computing devices, we propose here an approach that, to the best of our knowledge, has not been proposed so far: we want to examine the models of computation proposed in theoretical computer science to think about interactive computing systems. We offer a~literature survey where one can find \textit{explicit theories of interaction}.\footnote{We insist on our two criteria: \textit{explicit theories of interaction} in \textit{theoretical computer science}. We have in mind the fact that other communities e.g., the engineering community on reactive systems, are related to our topic but they have not conceptualized \textit{interaction} as such.} We show that the formal modelling of interactive computing systems has been brought down to whether the new interaction models are reducible to Turing’s \emph{a-machines} \parencite{Turing1937}---we will refer to them as Turing Machines (TMs). 
Questioning the theoretical bounds of the Turing Machine in computer science when faced with the existence of interactive computing devices has been explored at least since Milner's work on communicating and mobile systems \parencite{Milner1993, Milner1999}. To the best of our knowledge, there are three areas where interaction models are framed as such. These areas are some works (i) on concurrency by Milner and his followers \parencite{Milner1999, Milner2006}, (ii) on Reactive Turing Machines \parencite{Andersen1997, Baeten2013, VanLeeuwen2001, VanLeeuwen2006}, and (iii) on interaction as a~new non-algorithmic computing paradigm \parencite{Goldin2006, Wegner1997, Wegner2003}. For each of the three identified models, we:

@@

Milner introduced the opposition between \textit{interactional} and \textit{computational} behaviour. Introducing the concept of interaction, Milner \parencite*{Milner1975, Milner1982, Milner1983} referred to concurrent message passing between agents. Milner's work coincided with Petri’s \parencite*{Petri1980} new model of concurrent processes, which generally intended to describe concurrency in information systems.\footnote{Concurrency theory emerged from Dick Karp's early work in the 1960s, grew with \parencite{Petri1980} and later work on transition systems \parencite{Nielsen1981, Glabbeek2004},
and has now developed into a~mature theory of reactive systems \parencite{Harel1985} with diverse network models \parencites[for an overview, see][]{Lee1998}{Lee2006_modelsofcomputation}.}
 To Milner, interaction is more \textit{expressive} than a~TM, but it still describes an \textit{effective} procedure. Milner did not assert \textit{equivalence} between an interactive model and a~TM, but he introduced the topic \parencite{Milner1999} and left it unanswered. 
Four main differences between old (computational) and new (interactional) computing are made striking by Milner. First, in Milner’s words, a~Turing Machine prescribes a~behaviour to be executed. In contrast, new computing requires the description of an information flow between several system components. Second, old computing is characterized by a~\textit{hierarchical design} when current practice involves \textit{heterarchical} phenomena in the computing system. Third, in new computing, the designer cannot predict when agents will be triggered or the overall behaviour of the computing system. Fourth, the user is not merely looking for an end result in new computing practice. There is more than a~mathematical function to evaluate, as it used to be in old computing. The user instead interacts with the system, and the look for an end result is replaced by continuing interaction. Having taken stock of the evolution of computing practice on the engineering side, Milner examines its consequences on the logic foundations of computing. The pi-calculus and his work on the equivalence with automata, known as bisimulation, achieved this reflection on interactive processes \parencite{Milner1993, Milner1999} with a~formalism.

@@

If we look for a~mechanistic explanation of computing, we need some elements to be unpacked beyond a~formal account to make sense of the quoted scenario above. For example, we need to account for how reading and writing on the ports are possible. It presupposes that the interactive computing system can wait, pause, and react depending on the arrival or absence of new data. What allows such behavior? It presupposes some mechanisms allowing the system either to be interrupted by environmental processes or to check the new incoming values steadily.\footnote{More on these mechanisms and on the limitations of oracles can be found in \parencite{Martin2023}.} 
In other words, given the initial question (``what is an interactive computer?''), some phenomena cannot be accounted for within the frame of an extended Turing machine. The way oracles work remains at a~level of abstraction too remote from the minimal causal blueprint we need for our purpose.

@@

\begin{table}[]
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}llll@{}}
\rowcolor[HTML]{CBCEFB} 
%PP \toprule
\multicolumn{1}{c}{} &
  \multicolumn{1}{c}{\textbf{\begin{tabular}[c]{@{}c@{}}\large Interaction as concurrent \\ \large communicating systems\end{tabular}}} &
  \multicolumn{1}{c}{\textbf{\begin{tabular}[c]{@{}c@{}}\large Interaction as extended\\ \large Turing Machines\end{tabular}}} &
  \multicolumn{1}{c}{\textbf{\begin{tabular}[c]{@{}c@{}}\large Interaction as a~new\\ \large paradigm\end{tabular}}} \\ 
%PP  \midrule
\rowcolor[HTML]{FFFFFF} 
\textbf{\large Motivation} &
  \begin{tabular}[c]{@{}l@{}}Provides new logical\\ foundations to fit new\\ engineering challenges,\\ especially concurrency\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Extends the TM\\ model to account for\\ interactive devices\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Debunks the strong Church-\\ Turing thesis\\ \\ Discusses the scope of \\ algorithmic solving\\ \\ Prones the need for a~new\\ computing paradigm\end{tabular} \\
\rowcolor[HTML]{ECF4FF} 
\textbf{\begin{tabular}[c]{@{}l@{}}\large Account for\\ \large interaction\end{tabular}} &
  \begin{tabular}[c]{@{}l@{}}Information flow\\ \\ Heterarchical design\\ \\ No complete prediction about\\ overall behavior\\ \\ No end-result\\ \\ Process calculi\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}External data needed\\ during computation\\ \\ Non-uniformity of programs\\ \\ Interaction with agents\\ \\ Infinity of operations\\ \\ Interactive machines are TMs\\ with advice\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Computers have rich interaction\\ with the environment during\\ computing execution, \\ but this processing\\ is not merely algorithmic\end{tabular} \\
\rowcolor[HTML]{FFFFFF}  
\textbf{\begin{tabular}[c]{@{}l@{}}\large Uses and\\ \large criticisms\end{tabular}} &
  \begin{tabular}[c]{@{}l@{}}First conceptualization of \\ interaction\\ \\ Legacy for automata\\ theory\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Inspires the need for a~new\\ paradigm\\ \\ Puts at the forefront the\\ Church-Turing thesis\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Controversy about \\ the powerfulness of\\ the TM\end{tabular} \\
\rowcolor[HTML]{ECF4FF} 
\textbf{\begin{tabular}[c]{@{}l@{}} \large Issues for an account \\ \large of interactive computing\end{tabular}} &
  \begin{tabular}[c]{@{}l@{}}Definition of interaction\\ restricted to specific\\ properties: concurrency and\\ communication\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Formal oracles cannot\\ account for the physical \\ possibility of entering new\\ data\end{tabular} &
  \begin{tabular}[c]{@{}l@{}} Issues about powerfulness \\ and expressiveness constrict the\\ debate in the realm of\\ computability theory\end{tabular} \\ 
%PP \bottomrule
\end{tabular}%
}
\caption{Sum-up: an overview of explicit theories of interactive computing systems in theoretical computer science.}
\label{tab:my-table}
\end{table}

@@

But in a~narrow sense of expressiveness, one can interpret (wrongly, we think) the possibility of a~new paradigm as follows. Wegner and the tenants of Reactive Turing Machines could think of their interactive model as more expressive than a~TM, allowing their model to execute \textit{more functions}, even some of them being uncomputable functions in the sense of the Church-Turing thesis, solving the halting problem. In that case, the claim would indeed be controversial. The bold claim would be the following: a~TM is not only providing an account for algorithmic problem-solving through effective procedures but it could also be extended to account for other non-algorithmic processes, solving the uncomputable. Interaction would be some super-calculus, extending the calculative power of the original TM to account for interaction. It would be satisfactorily modeled with a~TM, only given more calculation power. It would go down the track of Accelerating Machines or Super-Turing Machines, able to calculate more than Turing's computable functions \parencite{Copeland2002, Copeland2011, MacLennan2009}.

@@

Such lessons have just started to be drawn. They have motivated, for example, distinctions between computational models and computational explanations \parencite{Klein2020} or between models of computation and computational mechanisms \parencite{Mikowski2014}. The lesson drawn is that formal models of computing systems do not provide us with the appropriate and complete level of description to build an explanation, which is expected to identify the relevant mechanisms at stake. More precisely, an explanation for computing phenomena requires bridging a~high-level description of a~computation and its blueprint \parencite{Mikowski2011, Mikowski2016}.
The approach is based on the standard of mechanistic explanation in science, coupled with the idea that a~computational process is intrinsically mechanistic:

\myquote{
Computational explanations, according to the mechanistic account are constitutive mechanistic explanations: they explain how a~mechanism's computational capacity is generated by the orchestrated operation of its component parts. To say that a~mechanism implements a~computation is to claim that the causal organization of the mechanism is such that the input and output information streams are causally linked and that this link, along with the specific structure of information processing, is completely described \parencite{Mikowski2014}.
}

@@

\myquote{
There are two ways in which computational models may correspond to mechanisms: first, they may be \textit{weakly equivalent} to the explanandum phenomenon, in that they only describe the input and output information, or \textit{strongly equivalent}, when they also correspond to the process that generates the output information \parencite{Mikowski2016}.
}

@@

\myquote{
Turing machines were not invented to be implemented physically at all, but some people still build them for fun. [\ldots] Imagine a~physical instantiation of a~trivial logical negation Turing machine, built of, say, steel and rubber and printing symbols on paper tape. Its alphabet of symbols consists of ``F'' and ``T.'' If the machine finds ``T'' on its tape, it rewrites it to ``F'' and halts; if it finds ``F,'' it rewrites it to ``T'' and halts. Let us suppose that the machine’s head is so old and worn out that it tears the paper tape during the readout. As a~result, no symbol will appear. [\ldots] Only when we describe the Turing machine literally, as a~causal system that has a~particular causal blueprint (engineering specifications of how it is built), can we causally predict such a~breakdown. [\ldots] Why are breakdowns and malfunctions so important? They help us discover the causal complexity of the system. [\ldots] an abstract model of computation will not predict all the possible outcomes of the breakdown, as it abstracts away from a~number of the system’s causal characteristics. So it will not tell us what is going to happen with the head; it will only say that the computation will no longer be correct \parencite{Mikowski2011}.
}

@@

We started from the need to update a~question in the philosophy of computing: \textit{what is a~computer}? Today's computers are highly interactive, so the question can be rephrased more precisely: \textit{what is an interactive computing system}? It is common to understand computers in terms of existing models of computation, hypothesizing that a~computer is primarily a~machine that carries out computation. Therefore, the working hypothesis has traditionally answered the initial question by asking what computation is. As already noted, this shift should not be taken for granted. There are, however, and the length of the paper does not allow it, historical and epistemological reasons for this shift that have been described and discussed \parencite{Daylight2014, DeMol2018a, Haigh2020}. We have chosen in this paper to ask ourselves if the shift is relevant in the case of interactive computing: do we understand what an interactive computer is by questioning the formal models proposed in theoretical computer science for interaction? Our literature review shows that there are better paths. There are two reasons for this. First, the conceptualization of interactive systems in theoretical computer science has focused on their comparison with the Turing machine (and sometimes other classical models), putting forward formal questions about powerfulness and equivalence of models that do not clarify the singularity of interactive systems from an epistemic point of view (rather than formal). There is an inherent difficulty in looking for an explanation of a~computing phenomenon in a~formal model: it needs more bricks to describe the mechanisms at stake, at a~level of abstraction operating the junction between a~high-level formalism and the blueprint. This work does not lead us to an aporia but to a~research program in the philosophy of computing: we must produce the right level of explanation for interactive computing.\footnote{More considerations on such a~research program are fleshed out in \parencite{Martin2023}.} This implies an identification of the mechanisms at play that make possible the interaction between processes within the computing machine (whether there are to be thought of as physical or computational processes, or a~mix of both\footnote{The distinction between computational and physical processes is out of the scope of this paper but more on this can be found e.g., in \parencite{Kycia_Niemczynowicz_2020}.}) and the environment. The components of such a~mechanism are to be identified and described at a~level of abstraction that allows a~satisfactory reference to the implementation.

### 2023-02-23 ART_Mscislawski/Mscislaw.tex
Editorial corrections added

Nevertheless, Krzanowski's concept of \textit{physical information} refers to the concept of information using the term \textit{concrete information} (IC). He refers to an extensive list of authors whose views converge in this respect
%\label{ref:RND0gu0V9TSmq}(e.g. Turek, 1978; von Weizsäcker, 1982; Nagel, 2012; Dodig-Crnkovic, 2013; Heller, 2014; Rovelli, 2016; Wilczek, 2016; Davies, 2020),
\parencites[e.g.,][]{turek_filozoficzne_1978}[von][]{von_weizsacker_einheit_1982}[][]{nagel_mind_2012}[][]{dodig-crnkovic_alan_2013}[][]{heller_elementy_2014}[][]{rovelli_meaning_2016}[][]{wilczek_beautiful_2016}[][]{davies_demon_2020}, %
 to name but a~few). He also seems to be guided by an opposition to attributing only the features of IA to information in general 
%\label{ref:RNDE2Cvw6yFhp}(cf. Krzanowski, 2020a; 2020c).
\parencites[cf.][]{krzanowski_does_2020}[][]{krzanowski_why_2020}. %
 This triggers a~need to introduce a~different approach to the concept of information, a~more qualitative one (IC). Thus, the fundamental features of IC can be described as follows 
%\label{ref:RND57FEnOikEz}(cf. Krzanowski, 2020b, p.2):
\parencite[cf.][p.2]{krzanowski_what_2020}:%
\begin{enumerate}[label=(IC\arabic*), start=0]
\item IC exists in space and time (i.e., spacetime) as a~physical object, which is why it is called \textit{concrete}.
\item With reference to IC0, IC is a~physical phenomenon, so it exists objectively and is not relative to anything.
\item IC has no intrinsic meaning.
\item IC is, in a~sense, responsible for the organization of the physical world.
\item IC's existence implies existence in the physical world, somewhere in the space-time continuum.
\end{enumerate}

>>>>

Nevertheless, Krzanowski's concept of \textit{physical information} refers to the concept of information using the term \textit{concrete information} (IC). He refers to an extensive list of authors whose views converge in this respect
%\label{ref:RND0gu0V9TSmq}(e.g. Turek, 1978; von Weizsäcker, 1982; Nagel, 2012; Dodig-Crnkovic, 2013; Heller, 2014; Rovelli, 2016; Wilczek, 2016; Davies, 2020),
\parencites[e.g.,][]{turek_filozoficzne_1978}[][]{von_weizsacker_einheit_1982}[][]{nagel_mind_2012}[][]{dodig-crnkovic_alan_2013}[][]{heller_elementy_2014}[][]{rovelli_meaning_2016}[][]{wilczek_beautiful_2016}[][]{davies_demon_2020}, %
 to name but a~few). He also seems to be guided by an opposition to attributing only the features of IA to information in general 
%\label{ref:RNDE2Cvw6yFhp}(cf. Krzanowski, 2020a; 2020c).
\parencites[cf.][]{krzanowski_does_2020}[][]{krzanowski_why_2020}. %
 This triggers a~need to introduce a~different approach to the concept of information, a~more qualitative one (IC). Thus, the fundamental features of IC can be described as follows 
%\label{ref:RND57FEnOikEz}(cf. Krzanowski, 2020b, p.2):
\parencite[cf.][p.2]{krzanowski_what_2020}:%
\begin{enumerate}[label=(IC\arabic*), start=0]
\item IC exists in space and time (i.e., spacetime) as a~physical object, which is why it is called \textit{concrete}.
\item With reference to IC0, IC is a~physical phenomenon, so it exists objectively and is not relative to anything.
\item IC has no intrinsic meaning.
\item IC is, in a~sense, responsible for the organization of the physical world.
\item IC's existence implies existence in the physical world, somewhere in the space-time continuum.
\end{enumerate}
